---
title: "Significance of Eplet Load in Liver Failure"
author: "Muhammad Sulman"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(corrplot)
library(survminer)
library(survival)
library(lubridate)
library(ggplot2)
library(ggfortify)
library(timereg)
library(car)
```

Function to load data from file and some reformatting.

```{r}
load_data <- function(filepath){
  #save csv to variable
  df <- read.csv(filepath)
  
  #subtract 1 from age at transplant
  df$Age.at.Transplant <- df$Age.at.Transplant - 1
  
  #add donor age
  df$Donor.DOB <- mdy(df$Donor.DOB)
  df$Date.Anesthesia.Start <- mdy(df$Date.Anesthesia.Start)
  df$Donor.Age <- as.numeric(df$Date.Anesthesia.Start - df$Donor.DOB, units="days")
  
  #Adds Time to Graft failure and status column
  df$Date.of.Graft.Failure <- mdy(df$Date.of.Graft.Failure)
  df$status <- ifelse(is.na(df$Date.of.Graft.Failure), 0 , 1)
  df$Time <- as.numeric(df$Date.of.Graft.Failure - df$Date.Anesthesia.Start, units="days")
  df <- df %>% filter(Time>0)
  
  #when the function is called, return the dataframe
  return(df)
}
```

Load data and save to a variable

```{r}
df <- load_data("data/liver_data.csv")
```

Saves the name of independent variable columns, including HLA and clinical factors, to a vector

```{r}
HLA_vars <- c("HLA.A.Epitope.Mismatch.No.",
              "HLA.B.Epitope.Mismatch.No.",
              "HLA.C.Epitope.Mismatch.No.",
              "HLA.DRB1.Epitope.Mismatch.No.",
              "HLA.DRB345.Epitope.Mismatch.No.",
              "HLA.DP.Epitope.Mismatch.No.",
              "HLA.DQ.Epitope.Mismatch.No.")

#Donor age is missing data, as such, may cause issues. Two variables created, one with and one without Donor age. 
clin_vars1 <- c("Number.of.Transplant",
                "BMI.of.Recipient",
                "Age.at.Transplant")

#with Donor Age
clin_vars2 <- c("Number.of.Transplant",
                "BMI.of.Recipient",
                "Age.at.Transplant", 
                "Donor.Age")

#all independant variables
indep_vars1 <- c(HLA_vars, clin_vars1)
indep_vars2 <- c(HLA_vars, clin_vars2)

#all variables, including Time, the dependant variable
all_vars1 <- c(indep_vars1, "Time", "status")
all_vars2 <- c(indep_vars2, "Time", "status")
```

Simplified data frames with relevant variables

```{r}
df_short1 <- df[all_vars1] %>% na.omit()
df_short2 <- df[all_vars2] %>% na.omit()
```

Make a correlation plot of all variables

```{r fig.width=12, fig.height=12}
#Note that age at transplant and Time negatively associate. DQ, DRB and DP also associate (likely proximal DNA regions, therefore lack of rearrangement) 
df_short2 %>% subset(select = -status) %>% cor() %>% corrplot()
```

Function for creating survival curves for every independant variable after stratifying by "high" (above median) and "low" (Below medium).

```{r}
plot_surv_and_logrank <- function(data, independent_vars, status_var, time_var) {
  results <- list()
  
  for (var in independent_vars) {
    data[[var]] <- ntile(data[[var]], 2) # Ensure data is modified by reference
    fit_formula <- as.formula(paste0("Surv(", time_var, ", ", status_var, ") ~ ", var))
    fit <- surv_fit(fit_formula, data = data)
    p <- ggsurvplot(
      fit, 
      data = data, 
      xlab = "Time", 
      ylab = "Survival probability", 
      palette = c("#4DAD5B","#FC4E07"),
      conf.int = FALSE,
      surv.scale = "percent",
      legend.labs = c(paste0("Low ", var), paste0("High ", var)) # Update legend labels
    )
    
    log_rank_result <- survdiff(fit_formula, data = data)

    # Store results
    results[[var]] <- list(plot = p, log_rank_result = log_rank_result)
    cat("Log-rank test result for", var, ":\n")
    print(log_rank_result)
    print(p)
    
    ggsave(paste0("img/", var, ".svg"), plot = p$plot)
  }

  return(results)
}
```

Call function and draw survival function for high (above median) and low of each independent variable. log rank tests are performed.

```{r}
#donor age, age at transplant, and number of transplant significantly reducve survival. 
result<-plot_surv_and_logrank(df_short2, indep_vars2, "status", "Time")
```

Function to output survival formula

```{r}
create_survival_formula <- function(independent_vars, status_var, time_var) {
  # Combine the status and time variables into the survival term of the formula
  survival_term <- paste0("Surv(", time_var, ", ", status_var, ")")

  independent_vars_str <- paste(independent_vars, collapse = " + ")
  formula_str <- paste(survival_term, "~", independent_vars_str)
  
  # Convert the string to an actual formula object
  formula_obj <- as.formula(formula_str)
  
  return(formula_obj)
}
```

CoxPH modelling

```{r fig.width=15, fig.height=15}
multi_form <- create_survival_formula(indep_vars2, "status", "Time")
cph_multi <- coxph(multi_form, data= df_short2)
summary(cph_multi)

#time varying effects of age, therefore it must be modeled differently
ggcoxzph(cox.zph(cph_multi))
```

Aalen's model was used, as previous residual analysis on cox model showed that there is a time-varying effect, making regular cox modelling inappropriate.


```{r}
df_short2$Donor.Age<-df_short2$Donor.Age/365

aalen_model <- aalen(Surv(Time, status) ~ HLA.A.Epitope.Mismatch.No. +
  HLA.B.Epitope.Mismatch.No. + HLA.C.Epitope.Mismatch.No. +
  HLA.DRB1.Epitope.Mismatch.No. +
  HLA.DRB345.Epitope.Mismatch.No. + HLA.DP.Epitope.Mismatch.No. + Donor.Age+
  HLA.DQ.Epitope.Mismatch.No.+ Number.of.Transplant + BMI.of.Recipient + Age.at.Transplant,
                             data = df_short2,max.time = 2000, n.sim = 5000, residuals = 1)

#test if model appropriate
average_dM <- apply(aalen_model$residuals$dM, 1, mean) # sum martingale increment
qqPlot(average_dM) #seems good enough, residuals are normally distributed
plot(aalen_model$residuals$time, average_dM, type = 'l', main = "Average Martingale Increments", xlab = "Time", ylab = "Average dM") #martingale residuals don't seem to have a relationship with time upon visual inspection


# Plot the average martingale increments against time
summary(aalen_model)
plot(aalen_model)
```


Age at transplant and number of transplant are significant, with age decreasing survival and number of prior transplants increasing survival. The hazard functions over time are shown above.

-   HLA-DP worsens survival and has a time varying effect .

-   Higher Age improves survival?! Maybe this is because older =\> weaker immune system =\> lower chance of rejection. Also time varying.

-   Number of transplant worsen survival. Time varying effect.

-   Chisq test pvalue indicates that the model has a significant fit to the data.
