---
title: "Mulinomial logistic regression of non-HLA data"
date: "2024-03-03"
author: "Muhammad Sulman"
email: "sulmanmu40@gmail.com"
github: "https://github.com/sulmanm1"
output: html_document
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r include=FALSE}
library(rmarkdown)
library(brant)
library(splines)
library(car)
library(tidyverse)
library(MASS)
library(brant)
library(effects)
library(corrplot)
library(jtools)
```

Function to load data from non-HLA. Converts all columns to the appropriate types.

```{r}
load_data <- function() {
  df<- read.csv("data/non_HLA.csv")
  df$Sex <- df$Sex %>% as.factor()
  df$Condition <- df$Condition %>% 
    as.factor()
  num_colnames<-c("AT1R.U.mL..CO.17.", "ENO1.Normalized..CO.4218.",
                  "FLRT2.Normalized..CO.688.", "VM.Normalized..CO.820.",
                  "TUBA1B.Normalized..CO.1987.", "IFIH1.Normalized..CO.3870.",
                  "PTPRN.Normalized..CO.3042.", "AURKA.Normalized..CO.4892.",
                  "PPIA.Normalized..CO.3292.", "EIF2A.Normalized..CO.6901.",
                  "PECR.Normalized..CO.4120.", "PRKCH.Normalized..CO.1048.",
                  "CXCL11.Normalized..CO.309.", "CXCL10.Normalized..CO.285.",
                  "AGRIN.Normalized..CO.504.", "GAPDH.Normalized..CO.508.",
                  "MYOSIN.Normalized..CO.9341.", "AGT.Normalized..CO.1641.",
                  "CHAF1B.Normalized..CO.11722.", "PLA2R.Normalized..CO.195.",
                  "LG3.Normalized..CO.3801.", "GSTT1.Normalized..CO.6136.",
                  "LMNA.Normalized..CO.6633.", "PRKCZ.Normalized..CO.9104.",
                  "LMNB1.Normalized..CO.2065.", "ARHGDIB.Normalized..CO.3918.",
                  "HNRNPK.Normalized..CO.845.", "IFNG.Normalized..CO.498.",
                  "REG3A.Normalized..CO.86.", "NUCLEOLIN.Normalized..CO.3669.",
                  "CD36.Normalized..CO.1591.", "TNFA.Normalized..CO.5331.",
                  "CXCL9.Normalized..CO.575.", "GDNF.Normalized..CO.1004.",
                  "COLLAGEN.I.Normalized..CO.375.", "COLLAGEN.II.Normalized..CO.155.",
                  "COLLAGEN.III.Normalized..CO.128.", "COLLAGEN.IV.Normalized..CO.71.",
                  "COLLAGEN.V.Normalized..CO.187.", "FIBRONECTIN.Normalized..CO.141.",
                  "Age.at.Transplant", "DSA_num")
  df <- df %>% mutate(across(all_of(num_colnames), ~as.numeric(as.character(.))))
  df$AGRIN.Normalized..CO.504. <- df$AGRIN.Normalized..CO.504./10
  df$AGT.Normalized..CO.1641. <- df$AGT.Normalized..CO.1641./100
  df$ARHGDIB.Normalized..CO.3918. <- df$ARHGDIB.Normalized..CO.3918./100
  df$AURKA.Normalized..CO.4892. <- df$AURKA.Normalized..CO.4892./100
  df$CD36.Normalized..CO.1591. <- df$CD36.Normalized..CO.1591./100
  df$CHAF1B.Normalized..CO.11722. <- df$CHAF1B.Normalized..CO.11722./100
  df$COLLAGEN.I.Normalized..CO.375. <- df$COLLAGEN.I.Normalized..CO.375./10
  df$COLLAGEN.II.Normalized..CO.155. <- df$COLLAGEN.II.Normalized..CO.155./10
  df$COLLAGEN.III.Normalized..CO.128. <- df$COLLAGEN.III.Normalized..CO.128./10
  df$COLLAGEN.IV.Normalized..CO.71. <- df$COLLAGEN.IV.Normalized..CO.71./10
  df$COLLAGEN.V.Normalized..CO.187. <- df$COLLAGEN.V.Normalized..CO.187./10
  df$CXCL10.Normalized..CO.285. <- df$CXCL10.Normalized..CO.285./10
  df$CXCL11.Normalized..CO.309. <- df$CXCL11.Normalized..CO.309./10
  df$CXCL9.Normalized..CO.575. <- df$CXCL9.Normalized..CO.575./100
  df$EIF2A.Normalized..CO.6901. <- df$EIF2A.Normalized..CO.6901./100
  df$ENO1.Normalized..CO.4218. <- df$ENO1.Normalized..CO.4218./100
  df$FIBRONECTIN.Normalized..CO.141. <- df$FIBRONECTIN.Normalized..CO.141./10
  df$FLRT2.Normalized..CO.688. <- df$FLRT2.Normalized..CO.688./100
  df$GAPDH.Normalized..CO.508. <- df$GAPDH.Normalized..CO.508./100
  df$GDNF.Normalized..CO.1004. <- df$GDNF.Normalized..CO.1004./100
  df$GSTT1.Normalized..CO.6136. <- df$GSTT1.Normalized..CO.6136./100
  df$HNRNPK.Normalized..CO.845. <- df$HNRNPK.Normalized..CO.845./100
  df$IFNG.Normalized..CO.498. <- df$IFNG.Normalized..CO.498./100
  df$IFIH1.Normalized..CO.3870. <- df$IFIH1.Normalized..CO.3870./100
  df$LG3.Normalized..CO.3801. <- df$LG3.Normalized..CO.3801./100
  df$LMNA.Normalized..CO.6633. <- df$LMNA.Normalized..CO.6633./100
  df$LMNB1.Normalized..CO.2065. <- df$LMNB1.Normalized..CO.2065./100
  df$MYOSIN.Normalized..CO.9341. <- df$MYOSIN.Normalized..CO.9341./100
  df$NUCLEOLIN.Normalized..CO.3669. <- df$NUCLEOLIN.Normalized..CO.3669./10
  df$PECR.Normalized..CO.4120. <- df$PECR.Normalized..CO.4120./100
  df$PLA2R.Normalized..CO.195. <- df$PLA2R.Normalized..CO.195./100
  df$PPIA.Normalized..CO.3292. <- df$PPIA.Normalized..CO.3292./100
  df$PRKCZ.Normalized..CO.9104. <- df$PRKCZ.Normalized..CO.9104./100
  df$PRKCH.Normalized..CO.1048. <- df$PRKCH.Normalized..CO.1048./100
  df$PTPRN.Normalized..CO.3042. <- df$PTPRN.Normalized..CO.3042./100
  df$REG3A.Normalized..CO.86. <- df$REG3A.Normalized..CO.86./10
  df$TNFA.Normalized..CO.5331. <- df$TNFA.Normalized..CO.5331./100
  df$TUBA1B.Normalized..CO.1987. <- df$TUBA1B.Normalized..CO.1987./100
  df$VM.Normalized..CO.820. <- df$VM.Normalized..CO.820./100
  return(df)
}
```

Function to replace normalized values with "1" for above cutoff and "0" for below. Cutoff numbers are in the column names.

```{r}
process_columns_with_cutoff <- function(colnames_vector, df) { 
  # Loop through each column name provided in the vector 
  for (colname in colnames_vector) {
    cutoff_str <- sub(".*CO\\.", "", colname)
    cutoff_str <- gsub("\\..*", "", cutoff_str)
    cutoff <- as.numeric(cutoff_str)
    df[[colname]] <- ifelse(df[[colname]] > cutoff, 1, 0)
    #df[[colname]] <- as.factor(df[[colname]])
  }
  return(df)
}

```

Function to plot distributions of continuous variables.

```{r}
plot_variable_distributions <- function(data, variable_names, outcome_var) { 
  # Pivot the data frame to long format 
  data_long <- pivot_longer(data, cols = all_of(variable_names), names_to = "variable", values_to = "value")

  # Filter out rows with NA values in 'value' to avoid issues in plotting
  data_long <- na.omit(data_long)
  
  # Plot data with aes() adjusted to include color based on the outcome variable
  p <- ggplot(data_long, aes(x = value, fill = as.factor(get(outcome_var)), color = as.factor(get(outcome_var)))) +
    geom_density(alpha = 0.3) +  # Adjust alpha for fill transparency
    facet_wrap(~variable, scales = "free", nrow = 20) +
    scale_fill_brewer(palette = "Set1", name = outcome_var) +
    scale_color_brewer(palette = "Set1", name = outcome_var) +  # Add this line to use consistent color palette for lines
    theme_minimal() +
    theme(legend.position = "bottom") +
    labs(x = "Value", y = "Density") 
  
  # Alter text for readability
  p <- p + theme(strip.text = element_text(size = 8),  # Text size
                 strip.text.x = element_text(angle = 0, hjust = 0.5),  # Adjust angle and horizontal justification
                 strip.background = element_rect(fill = "white"))  # Change background color for visibility
  print(p)
  ggsave("variable_distributions.svg")
}
```

Function that returns subsetted data frame based on the column names provided

```{r}
create_subset_df <- function(original_df, colnames_array) {
  # Ensure that the column names exist in the original dataframe 
  valid_colnames <- colnames_array[colnames_array %in% names(original_df)]

  subset_df <- original_df[, valid_colnames, drop = FALSE]
  return(subset_df)
}
```

Return all highly correlated variables

```{r}
find_highly_correlated_variables <- function(dataframe, column_names, threshold = 0.8) {
  subset_df <- dataframe[, column_names]
  corr_matrix <- cor(subset_df, use = "complete.obs")  # 'complete.obs' handles missing values by using complete cases
  high_corr_pairs <- which(abs(corr_matrix) > threshold & lower.tri(corr_matrix), arr.ind = TRUE)
  
  # Extract the variable names for these pairs
  high_corr_vars <- unique(c(column_names[high_corr_pairs[, "row"]], column_names[high_corr_pairs[, "col"]]))
  
  return(high_corr_vars)
}
```

Write formulas for the multinomial logistic regression

```{r}
create_pom_formula <- function(response_var, clin_cofactors, independent_vars, df) {
  all_vars <- c(response_var, clin_cofactors, independent_vars) 
  missing_vars <- setdiff(all_vars, names(df))
  
  if(length(missing_vars) > 0) { stop(paste("The following variables are missing in the dataframe:",
                                             paste(missing_vars, collapse=", "), ".")) 
    }
  predictors <- c(clin_cofactors, independent_vars)
  formula_str <- paste(response_var, "~", paste(predictors, collapse = " + "))
  return(as.formula(formula_str))
}
```

Function to create a PCA column

```{r}
addPCAComponent <- function(dataframe, columns, pcaColumnName) {
  subset_df <- dataframe[, columns, drop = FALSE]
  pca_result <- prcomp(subset_df, center = TRUE, scale. = TRUE)
  
  # Extract the first principal component
  first_pca_component <- pca_result$x[, 1]
  dataframe[[pcaColumnName]] <- first_pca_component
  return(dataframe)
}
```

Saving the names of the independent variables and the continuous independent variables

```{r}
#missing fibronectin and pla2r because they have low levels of fluorescence => inaccurate OR.
tot_nHLA_var<-c("AT1R.U.mL..CO.17.",
"ENO1.Normalized..CO.4218.", "FLRT2.Normalized..CO.688.",
"VM.Normalized..CO.820.", "TUBA1B.Normalized..CO.1987.",
"IFIH1.Normalized..CO.3870.", "PTPRN.Normalized..CO.3042.",
"AURKA.Normalized..CO.4892.", "PPIA.Normalized..CO.3292.",
"EIF2A.Normalized..CO.6901.", "PECR.Normalized..CO.4120.",
"PRKCH.Normalized..CO.1048.", "CXCL11.Normalized..CO.309.",
"CXCL10.Normalized..CO.285.", "AGRIN.Normalized..CO.504.",
"GAPDH.Normalized..CO.508.", "MYOSIN.Normalized..CO.9341.",
"AGT.Normalized..CO.1641.", "CHAF1B.Normalized..CO.11722.",
"LG3.Normalized..CO.3801.", "GSTT1.Normalized..CO.6136.", "LMNA.Normalized..CO.6633.",
"PRKCZ.Normalized..CO.9104.", "LMNB1.Normalized..CO.2065.",
"ARHGDIB.Normalized..CO.3918.", "HNRNPK.Normalized..CO.845.",
"IFNG.Normalized..CO.498.", "REG3A.Normalized..CO.86.",
"NUCLEOLIN.Normalized..CO.3669.", "CD36.Normalized..CO.1591.",
"TNFA.Normalized..CO.5331.", "CXCL9.Normalized..CO.575.",
"GDNF.Normalized..CO.1004.", "COLLAGEN.I.Normalized..CO.375.",
"COLLAGEN.II.Normalized..CO.155.", "COLLAGEN.III.Normalized..CO.128.",
"COLLAGEN.IV.Normalized..CO.71.", "COLLAGEN.V.Normalized..CO.187.")

tot_nHLA_var2<-c("AT1R.U.mL..CO.17.",
"ENO1.Normalized..CO.4218.", "FLRT2.Normalized..CO.688.",
"VM.Normalized..CO.820.", "TUBA1B.Normalized..CO.1987.",
"IFIH1.Normalized..CO.3870.", "PTPRN.Normalized..CO.3042.",
"AURKA.Normalized..CO.4892.", "PPIA.Normalized..CO.3292.",
"EIF2A.Normalized..CO.6901.", "PECR.Normalized..CO.4120.",
"PRKCH.Normalized..CO.1048.", "CXCL11.Normalized..CO.309.",
"CXCL10.Normalized..CO.285.", "AGRIN.Normalized..CO.504.",
"GAPDH.Normalized..CO.508.", "MYOSIN.Normalized..CO.9341.",
"AGT.Normalized..CO.1641.", "CHAF1B.Normalized..CO.11722.",
"PLA2R.Normalized..CO.195.", "LG3.Normalized..CO.3801.",
"GSTT1.Normalized..CO.6136.", "LMNA.Normalized..CO.6633.",
"PRKCZ.Normalized..CO.9104.", "LMNB1.Normalized..CO.2065.",
"ARHGDIB.Normalized..CO.3918.", "HNRNPK.Normalized..CO.845.",
"IFNG.Normalized..CO.498.", "REG3A.Normalized..CO.86.",
"NUCLEOLIN.Normalized..CO.3669.", "CD36.Normalized..CO.1591.",
"TNFA.Normalized..CO.5331.", "CXCL9.Normalized..CO.575.",
"GDNF.Normalized..CO.1004.", "COLLAGEN.I.Normalized..CO.375.",
"COLLAGEN.II.Normalized..CO.155.", "COLLAGEN.III.Normalized..CO.128.",
"COLLAGEN.IV.Normalized..CO.71.", "COLLAGEN.V.Normalized..CO.187.",
"FIBRONECTIN.Normalized..CO.141.")

tot_clin_var<-c("Age.at.Transplant", "DSA_num")

tot_independant_var_cont<-c(tot_nHLA_var, tot_clin_var)

nc_var<-c("AT1R.U.mL..CO.17.", "ENO1.Normalized..CO.4218.",
"FLRT2.Normalized..CO.688.", "IFIH1.Normalized..CO.3870.",
"AURKA.Normalized..CO.4892.", "PPIA.Normalized..CO.3292.",
"EIF2A.Normalized..CO.6901.", "PECR.Normalized..CO.4120.",
"PRKCH.Normalized..CO.1048.", "CXCL11.Normalized..CO.309.",
"CXCL10.Normalized..CO.285.", "AGRIN.Normalized..CO.504.",
"GAPDH.Normalized..CO.508.", "MYOSIN.Normalized..CO.9341.",
"CHAF1B.Normalized..CO.11722.", "PLA2R.Normalized..CO.195.",
"LG3.Normalized..CO.3801.", "GSTT1.Normalized..CO.6136.",
"LMNA.Normalized..CO.6633.", "PRKCZ.Normalized..CO.9104.",
"LMNB1.Normalized..CO.2065.", "ARHGDIB.Normalized..CO.3918.",
"HNRNPK.Normalized..CO.845.", "IFNG.Normalized..CO.498.",
"REG3A.Normalized..CO.86.", "NUCLEOLIN.Normalized..CO.3669.",
"TNFA.Normalized..CO.5331.", "CXCL9.Normalized..CO.575.",
"GDNF.Normalized..CO.1004.", "COLLAGEN.I.Normalized..CO.375.",
"COLLAGEN.II.Normalized..CO.155.", "COLLAGEN.III.Normalized..CO.128.",
"COLLAGEN.IV.Normalized..CO.71.", "COLLAGEN.V.Normalized..CO.187.",
"FIBRONECTIN.Normalized..CO.141.")
```

Replace normalized fluorescence values with "1", presence or "0" absence.

```{r}
df_nocut<- load_data()
#not doing because screws analysis
#df<-process_columns_with_cutoff(tot_nHLA_var, df_nocut)
df<-df_nocut

```

Plot the distributions of continuous variables

```{r fig.width=12, fig.height=40}
plot_variable_distributions(df_nocut, tot_independant_var_cont, "Condition")
```

Subset into pre and post transplant data

```{r}
df_pre<- subset(df, Pre==1)
df_post<- subset(df, Pre==0)
```

Correlation plot for pre-transplant data

```{r fig.width=12, fig.height=12}
df_pre_nc <- df_nocut %>% filter(Pre==1)
create_subset_df(df_pre_nc, tot_independant_var_cont) %>% cor() %>% corrplot()
find_highly_correlated_variables(df_pre_nc, tot_independant_var_cont, 0.8)
```

Correlation plot for post-transplant data

```{r fig.width=12, fig.height=12}
df_post_nc <- df_nocut %>% filter(Pre==0)
create_subset_df(df_post_nc, tot_independant_var_cont) %>% cor() %>% corrplot()
find_highly_correlated_variables(df_post_nc, tot_independant_var_cont, 0.8)
```

variables that are colinear are replaced with a PCA

```{r}
all_colin<-c("VM.Normalized..CO.820.", "TUBA1B.Normalized..CO.1987.", "AGT.Normalized..CO.1641.", "CD36.Normalized..CO.1591.", "PPIA.Normalized..CO.3292.", "COLLAGEN.III.Normalized..CO.128.", "COLLAGEN.IV.Normalized..CO.71.", "FLRT2.Normalized..CO.688.", "COLLAGEN.I.Normalized..CO.375.")

colin_set1<-c("VM.Normalized..CO.820.", "TUBA1B.Normalized..CO.1987.", "AGT.Normalized..CO.1641.", "CD36.Normalized..CO.1591.", "PPIA.Normalized..CO.3292.", "FLRT2.Normalized..CO.688.")
colin_set2<-all_colin[!all_colin %in% colin_set1]

df_pre<-addPCAComponent(df_pre, colin_set1, "PCA1")
df_post<-addPCAComponent(df_post, colin_set1, "PCA1")

df_pre<-addPCAComponent(df_pre, colin_set2, "PCA2")
df_post<-addPCAComponent(df_post, colin_set2, "PCA2")

df_pre<-df_pre %>% dplyr::select(-all_of(all_colin))
df_post<-df_post %>% dplyr::select(-all_of(all_colin))

tot_nHLA_var<-tot_nHLA_var[!tot_nHLA_var %in% all_colin]
tot_nHLA_var<-c(tot_nHLA_var, "PCA1", "PCA2")
tot_independant_var_cont<-tot_independant_var_cont[!tot_independant_var_cont %in% all_colin]
tot_independant_var_cont<-c(tot_independant_var_cont, "PCA1", "PCA2")

```


Function to remove Columns in which there is only one instance (only presence or only absence) as this does not help with prediction.
```{r}
find_factor_columns_with_missing_level <- function(dataframe, columns) {
  # Initialize an empty vector to store the names of columns that meet the criteria
  columns_with_missing_level <- c()
  
  for (col_name in columns) {
    if (is.factor(dataframe[[col_name]]) && length(levels(dataframe[[col_name]])) == 2) {
      level_counts <- table(dataframe[[col_name]])
      
      # Check if any of the levels has no instances
      if (any(level_counts == 0)) {
        columns_with_missing_level <- c(columns_with_missing_level, col_name)
      }
    }
  }
  return(columns_with_missing_level)
}

```

Create datasets for DGF and failure analysis

```{r}
df_pre_dgf<- subset(df_pre, !Condition==2)
df_pre_dgf$Condition <- factor(df_pre_dgf$Condition)

df_post_dgf<-subset(df_post, !Condition==2)
df_post_dgf$Condition <- factor(df_post_dgf$Condition)

df_pre_fail<-subset(df_pre, !Condition==1)
df_pre_fail$Condition <- factor(df_pre_fail$Condition)

df_post_fail<-subset(df_post, !Condition==1)
df_post_fail$Condition <- factor(df_post_fail$Condition)

```

Multinomial logistic regression for pre-transplant data

```{r}
pre_dgf_formula<-create_pom_formula("Condition", tot_clin_var, tot_nHLA_var, df_pre)
pre_dgf_model<-glm(pre_dgf_formula, family = "binomial",data = df_pre_dgf)
#pre_dgf_step.model<-stepAIC(pre_dgf_model, direction = "backward", trace = FALSE)
#optireg_pre_dgf<-pre_dgf_step.model$terms
#pre_dgf_model_opt<-glm(optireg_pre_dgf, family = "binomial",data = df_pre_dgf)
summary(pre_dgf_model)

pre_fail_formula<-create_pom_formula("Condition", tot_clin_var, tot_nHLA_var, df_pre)
pre_fail_model<-glm(pre_fail_formula, family = "binomial",data = df_pre_fail)
#pre_fail_step.model<-stepAIC(pre_fail_model, direction = "backward", trace = FALSE)
#optireg_pre_fail<-pre_fail_step.model$terms
#pre_fail_model_opt<-glm(optireg_pre_fail, family = "binomial",data = df_pre_fail)
summary(pre_fail_model)

```

Multinomial logistic regression for post-transplant data

```{r}
post_dgf_formula<-create_pom_formula("Condition", tot_clin_var, tot_nHLA_var, df_post)
post_dgf_model<-glm(post_dgf_formula, family = "binomial",data = df_post_dgf)
#post_dgf_step.model<-stepAIC(post_dgf_model, direction = "backward", trace = FALSE)
#optireg_post_dgf<-post_dgf_step.model$terms
#post_dgf_model_opt<-glm(optireg_post_dgf, family = "binomial",data = df_post_dgf)
summary(post_dgf_model)

post_fail_formula<-create_pom_formula("Condition", tot_clin_var, tot_nHLA_var, df_post)
post_fail_model<-glm(post_fail_formula, family = "binomial",data = df_post_fail)
#post_fail_step.model<-stepAIC(post_fail_model, direction = "backward", trace = FALSE)
#optireg_post_fail<-post_fail_step.model$terms
#post_fail_model_opt<-glm(optireg_post_fail, family = "binomial",data = df_post_fail)
summary(post_fail_model)
```

```{r echo=F}
library(sjPlot)
library(sjlabelled)
library(sjmisc)
```
Plotting the Odds Ratios

```{r}
library(forestmodel)

to_rm<-c("COLLAGEN.V.Normalized..CO.187.", "PCA1", "PCA2", "FIBRONECTIN.Normalized..CO.141.", "PLA2R.Normalized..CO.195.", "DSA_num", "Age.at.Transplant")

to_include<-tot_independant_var_cont[!tot_independant_var_cont %in% to_rm]
f_pre<-forest_model(model_list=list("Delayed Graft function"=pre_dgf_model, "Organ Rejection"=pre_fail_model), covariates = to_include, format_options = forest_model_format_options(point_size = 2))
f_pre
ggsave("pre_trans_plot.svg")

f_post<-forest_model(model_list=list("Delayed Graft function"=post_dgf_model, "Organ Rejection"=post_fail_model), covariates = to_include, format_options = forest_model_format_options(point_size =2))
f_post
ggsave("post_trans_plot.svg")

```

```{}